A multi-dimensional extension of the Lightweight Temporal Compression method

### 2018-10-17 paper.tex
Now fits in 6 pages

\begin{figure}
\includegraphics[width=\columnwidth]{ltc.pdf}
\caption{Illustration of the LTC algorithm and notations used. Blue 
dots are received points, red dots are transmitted points. Dashed lines 
represent the high line and the low line at times when a point is 
transmitted.\vspace*{-0.3cm}}
\label{fig:ltc}
\end{figure}
\todo{show $\xi$ in the Figure}

@@

LTC approximates the data stream
by a piece-wise linear function of time, ensuring that the error at
every time-point is bounded by parameter $\epsilon$.

@@

Let $(u_0^k, y_0^k) \in \mathbb{R}^{n+1}$ be the latest transmitted point. For convenience, all the subsequent points will be
expressed in the orthogonal space with origin $(\tau_k, \xi_k)$. We denote by $(v_j, z_j)_{j \in \llbracket 0, m \rrbracket}$ such points:
\begin{equation*}
\forall j \leq m,\  (v_j, z_j) = (u_j^k - \tau_k, y_j^k - \xi_k)
\end{equation*}
We define $\mathcal{B}_j$ as the ball of $\mathbb{R}^n$ of center $\frac{v_1}{v_j}z_j$ and of radius
$\frac{v_1}{v_j}\epsilon$:
\begin{equation*}
\mathcal{B}_j = \left\{ z \in \mathbb{R}^n,\  \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon \right\}
\end{equation*}

@@

We formulate the \emph{LTC property} as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{\frac{v_j}{v_1}z-z_j} \leq \epsilon.
\end{equation*}
The original LTC algorithm ensures that the LTC property is
verified between each transmission. Indeed, all the data points
$z$ such that $(v_1, z)$ is between the high line and the low line
verify the property. Line 9 in Algorithm~\ref{algo:ltc} guarantees that
such a point exists.

@@

The only feasible algorithm that we found is norm-specific. It
maintains a representation of the intersection
$\bigcap_{j=1}^{m}{\mathcal{B}_j}$ which is updated at every iteration.
The intersection tests can then be done in constant time. However,
updating the representation of the intersection may be costly
depending on the norm used. For the infinity norm, the representation
is a rectangular cuboid which is straightforward to update by
intersection with an n-ball, since the n-ball is a rectangular cuboid too.
For the Euclidean norm, the representation is a volume with no particular property,
which is more costly to maintain.

@@

As mentioned before, norm selection in $\mathbb{R}^n$ has a critical
impact on the compression error and ratio. To appreciate this effect,
let us consider the comparison between the infinity norm and the
Euclidean norm in dimension 2. By comparing the unit disk to a
square of side 2, we obtain that the compression ratio of a random stream would
be $\frac{4}{\pi}$ times larger with the infinity norm than with the 
Euclidean norm. In 3D, this ratio would be $\frac{6}{\pi}$. Conversely, 
a compression error bounded by $\epsilon$ with the infinity norm 
corresponds to a compression error of $\frac{\epsilon}{\sqrt{n}}$ with 
the Euclidean norm. Unsurprisingly, the
infinity norm is more 'tolerant' than the Euclidean norm.

It should also be noted that using the infinity norm in dimension $n$ 
boils down to the use of the 1D LTC algorithm independently in each 
dimension, since a data point will be transmitted as soon as the linear 
approximation doesn't hold in any of the dimensions.

@@

To implement LTC with the infinity norm, we maintain a cuboid 
representation of $\cap_{l=1}^j{\mathcal{B}_l}$ throughout the 
iterations of the \texttt{while} loop in 
Algorithm~\ref{algo:general-ltc}. The implementation only requires 
constant memory and reduced CPU time to update the cuboid with the 
n-ball associated with the new data point at each iteration.

With the Euclidean norm, the intersection test is more complex, as 
mentioned previously. We keep in memory a growing set $S$ of balls and 
the bounding box $B$ of their intersection. Then, when a new point 
arrives, we consider the associated ball $\mathcal{B}_j$ and our 
intersection test works as in Algorithm~\ref{algo:euclidean}. 
\texttt{box} is a function that returns the bounding box of an n-ball. 
\texttt{find\_bisection(S, B)} searches for a point in all the elements 
in S, using plane sweep and bisection initialized by the bounds of B.

@@

Our code is available at 
\url{https://github.com/big-data-lab-team/stream-summarization} under
MIT license.

@@

In this experiment, we validated the behavior of our LTC extension on a 
PC using data acquired with Neblina. We collected two 3D 
accelerometer time series, a short one and a longer one, acquired 
on two different subjects performing biceps curl, with a 50Hz sampling rate (see 
Figure~\ref{fig:datasets-1}). In both cases, the 
subject was wearing Neblina on their right wrist. It should be noted that the longer dataset also has 
a higher amplitude, most likely due to differences between subjects.

@@

\paragraph{Infinity vs Euclidean norms}
Compared to the infinity norm, the average ratio between the compression ratios obtained
with the infinity and Euclidean norm is 1.08 for 2D data, and 1.20
for 3D data. These ratios are lower than the theoretical values of
$\frac{4}{\pi}$ in 2D and $\frac{6}{\pi}$ in 3D, which are obtained for
random-uniform signals. Unsurprisingly, the infinity norm surpasses the
Euclidean norm in terms of resource consumption. Memory-wise, the
infinity norm requires a constant amount of 80~B, used to store the
intersection of n-balls. The Euclidean norm, however, uses up to 4.9~KB of memory
for the Long time series in 3D with $\epsilon$=48.8~mg. More importantly,
the amount of required memory increases for more complex time series
(Long vs Short), and it also increases with larger
values of $\epsilon$. Similar observations are made for the processing
time, with values ranging from 0.2~ms for the simplest time series and
smallest $\epsilon$, to 26.4~ms for the most complex time series and
largest $\epsilon$.
%~ Figure~\ref{fig:memory} shows the memory consumption 
%~ of the 3D Euclidean implementation for $\epsilon$=48.8~mg: peaks appear
%~ at the end of long compressed ranges where the signal was $\epsilon$-closed
%~ to the linear approximation.

\paragraph{2D vs 3D}
For a given $\epsilon$ of 48.8~mg and a given time series, the compression 
ratios are always higher in 2D than in 3D. It makes sense since the
probability for the signal to deviate from a straight line
approximation is higher in 3D than it is in 2D. Besides, resource
consumption is higher in 3D than in 2D: for the infinity norm, 3D
consumes 1.4 times more memory than 2D (1.6 times on average for
Euclidean norm), and processing time is 1.3 longer on average (1.18 on
average for Euclidean norm).

@@

\begin{figure*}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-x.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-y.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-z.pdf}
\caption{Walking}
\end{subfigure}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-x.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-y.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-z.pdf}
\caption{Running}
\end{subfigure}
\caption{Time series used in Experiment 2}
\label{fig:datasets-2}
\end{figure*}

@@

We acquired two 3D accelerometer time series at 200~Hz corresponding to
common human activities: walking and running (see Figure~\ref{fig:datasets-2}). In both cases, the
subject was wearing Neblina on their right wrist.
We collected 1,000 data points for each activity, corresponding to
5 seconds of activity.

We measured energy consumption associated with the transmission of
these time series, for different norms and $\epsilon$ values. To do so,
we `replayed' the time series by loading them as a byte array in
Neblina, and we measured the current every 500~ms.

@@

\section*{Acknowledgement}
This work was funded by a Strategic Project Grant of the Natural 
Sciences and Engineering Research Council of Canada. We thank Hosein 
Nourani for providing the data used in Experiment 1.

>>>>

\begin{figure}
\includegraphics[width=\columnwidth]{ltc.pdf}
\caption{Illustration of the LTC algorithm. Blue 
dots are received points, red dots are transmitted points. Dashed lines 
represent the high and low lines when a point is 
transmitted.\vspace*{-0.3cm}}
\label{fig:ltc}
\end{figure}

\todo{show $\xi$ in the Figure}
\todo{Fix Omid's comment in the Figure}

@@

LTC approximates the data stream
by a piece-wise linear function of time, with an error bounded by parameter $\epsilon$.

@@

Let $(u_0^k, y_0^k) \in \mathbb{R}^{n+1}$ be the latest transmitted point. For convenience, all the subsequent points will be
expressed in the orthogonal space with origin $(\tau_k, \xi_k)$. We denote by $(v_j, z_j)_{j \in \llbracket 0, m \rrbracket}$ such points:
\begin{equation*}
\forall j \leq m,\  (v_j, z_j) = (u_j^k - \tau_k, y_j^k - \xi_k)
\end{equation*}
Let $\mathcal{B}_j$ be the ball of $\mathbb{R}^n$ of center $\frac{v_1}{v_j}z_j$ and radius
$\frac{v_1}{v_j}\epsilon$:
\begin{equation*}
\mathcal{B}_j = \left\{ z \in \mathbb{R}^n,\  \norm{z-\frac{v_1}{v_j}z_j} \leq \frac{v_1}{v_j}\epsilon \right\}
\end{equation*}

@@

We define the \emph{LTC property} as follows:
\begin{equation*}
\exists z \in \mathbb{R}^n, \ \forall j \in \llbracket 1, m \rrbracket, \norm{\frac{v_j}{v_1}z-z_j} \leq \epsilon.
\end{equation*}
The original LTC algorithm ensures that the LTC property is
verified between each transmission. Indeed, all the data points
$z$ such that $(v_1, z)$ is between the high line and the low line
verify the property. Line 9 in Algorithm~\ref{algo:ltc} guarantees that
such a point exists.

@@

The only feasible algorithm that we found is norm-specific. It
maintains a representation of the intersection
$\bigcap_{j=1}^{m}{\mathcal{B}_j}$ which is updated at every iteration.
The intersection tests can then be done in constant time. However,
updating the representation of the intersection may be costly
depending on the norm used. For the infinity norm, the representation
is a rectangular cuboid which is straightforward to update by
intersection with an n-ball.
For the Euclidean norm, the representation is a volume with no particular property,
which is more costly to maintain.

@@

As mentioned before, norm selection in $\mathbb{R}^n$ has a critical
impact on the compression error and ratio. To appreciate this effect,
let us compare the infinity norm and the
Euclidean norm in dimension 2. By comparing the unit disk to a
square of side 2, we obtain that the compression ratio of a random stream would
be $\frac{4}{\pi}$ times larger with the infinity norm than with the 
Euclidean norm. In 3D, this ratio would be $\frac{6}{\pi}$. Conversely, 
a compression error bounded by $\epsilon$ with the infinity norm 
corresponds to a compression error of $\frac{\epsilon}{\sqrt{n}}$ with 
the Euclidean norm. Unsurprisingly, the
infinity norm is more `tolerant' than the Euclidean norm.

It should also be noted that using the infinity norm in $\mathbb{R}^n$ 
boils down to the use of the 1D LTC algorithm independently in each 
dimension, since a data point will be transmitted as soon as the linear 
approximation doesn't hold in any of the dimensions.

@@

To implement LTC with the infinity norm, we maintain a cuboid 
representation of $\cap_{l=1}^j{\mathcal{B}_l}$ throughout the 
iterations of the \texttt{while} loop in 
Algorithm~\ref{algo:general-ltc}. The implementation works with
constant memory and requires limited CPU time.

With the Euclidean norm, the intersection test is more complex. We keep 
in memory a growing set $S$ of balls and the bounding box $B$ of their 
intersection. Then, when a new point arrives, we consider the 
associated ball $\mathcal{B}_j$ and our intersection test works as in 
Algorithm~\ref{algo:euclidean}. \texttt{box} is a function that returns 
the bounding box of an n-ball. \texttt{find\_bisection(S, B)} searches 
for a point in all the elements in S, using plane sweep and bisection 
initialized by the bounds of B. Our code is available at 
\url{https://github.com/big-data-lab-team/stream-summarization} under 
MIT license.

@@

@@

We validated the behavior of our LTC extension on a 
PC using data acquired with Neblina. We collected two 3D 
accelerometer time series, a short one and a longer one, acquired 
on two different subjects performing biceps curl, with a 50Hz sampling rate (see 
Figure~\ref{fig:datasets-1}). In both cases, the 
subject was wearing Neblina on their right wrist. It should be noted that the longer dataset also has 
a higher amplitude, most likely due to differences between subjects.

@@

\paragraph{Infinity vs Euclidean norms}
Compared to the infinity norm, the average ratio between the compression ratios obtained
with the infinity and Euclidean norm is 1.08 for 2D data, and 1.20
for 3D data. These ratios are lower than the theoretical values of
$\frac{4}{\pi}$ in 2D and $\frac{6}{\pi}$ in 3D, which are obtained for
random-uniform signals. Unsurprisingly, the infinity norm surpasses the
Euclidean norm in terms of resource consumption. Memory-wise, the
infinity norm requires a constant amount of 80~B, used to store the
intersection of n-balls. The Euclidean norm, however, uses up to 4.9~KB of memory
for the Long time series in 3D with $\epsilon$=48.8~mg. More importantly,
the amount of required memory increases for more complex time series
(Long vs Short), and it also increases with larger
values of $\epsilon$. Similar observations are made for the processing
time, with values ranging from 0.2~ms for the simplest time series and
smallest $\epsilon$, to 26.4~ms for the most complex time series and
largest~$\epsilon$.
%~ Figure~\ref{fig:memory} shows the memory consumption 
%~ of the 3D Euclidean implementation for $\epsilon$=48.8~mg: peaks appear
%~ at the end of long compressed ranges where the signal was $\epsilon$-closed
%~ to the linear approximation.

\paragraph{2D vs 3D}
For a given $\epsilon$, the compression 
ratios are always higher in 2D than in 3D. It makes sense since the
probability for the signal to deviate from a straight line
approximation is higher in 3D than it is in 2D. Besides, resource
consumption is higher in 3D than in 2D: for the infinity norm, 3D
consumes 1.4 times more memory than 2D (1.6 times on average for
Euclidean norm), and processing time is 1.3 longer (1.18 on
average for Euclidean norm).

@@

\begin{figure*}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-x.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-y.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/walking-1000-z.pdf}
\caption{Walking}
\end{subfigure}
\begin{subfigure}{2\columnwidth}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-x.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-y.pdf}
\includegraphics[width=0.3\columnwidth]{./figures/running-1000-z.pdf}
\caption{Running}
\end{subfigure}
\caption{Time series used in Experiment 2 \vspace*{-0.3cm}}
\label{fig:datasets-2}
\end{figure*}

@@

We acquired two 3D accelerometer time series at 200~Hz for two 
activities: walking and running (see Figure~\ref{fig:datasets-2}). In 
both cases, the subject was wearing Neblina on their right wrist. We 
collected 1,000 data points for each activity, corresponding to 5 
seconds of activity.

We measured energy consumption associated with the transmission of
these time series by
`replaying' the time series after loading them as a byte array in
Neblina. We measured the current every 500~ms.

@@

\section*{Acknowledgement}
This work was funded by the Natural 
Sciences and Engineering Research Council of Canada. We thank Hosein 
Nourani for providing data for Experiment 1.
